---
title: "Introduction to {ospsuite.bmlm}"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to {ospsuite.bmlm}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  error = FALSE,
  comment = "#>"
)
```

```{r setup, echo=FALSE,warning=FALSE,message=FALSE}
library(ospsuite.bmlm)
```

## What is {ospsuite.bmlm}?

The {ospsuite.bmlm} package provides tools for fitting Bayesian multilevel models within the OSPSuite framework. This package leverages the capabilities of the {ospsuite.reportingframework} package for project configuration, allowing users to efficiently set up their modeling environment.

Bayesian multilevel models are intended for parameter identification using individual study data. They are particularly useful for analyzing hierarchical data structures, where parameters of interest include:

- **Global parameters**: Such as compound properties (e.g., lipophilicity, molecular weight)
- **Distribution parameters**: Of individual parameters such as clearance or absorption rates
- **Individual parameters**: Patient-specific values that vary within the population

The modeling process employs a virtual twin population, where for each study individual, a virtual individual is generated based on anthropometric data. The fitting process uses maximum likelihood estimation (MLE), with optimization performed using the R function `optim`.

## Key Features

- **Integration with OSPSuite**: Seamless workflow integration with the reporting framework
- **Flexible Configuration**: Excel-based configuration files for easy setup
- **Multiple Optimization Methods**: Support for various optimization algorithms (BFGS, Nelder-Mead, etc.)
- **Internal Optimization**: Optional internal optimization for improved efficiency
- **Comprehensive Diagnostics**: Built-in plotting functions for monitoring optimization progress
- **Prior Distributions**: Support for various prior distributions on parameters
- **Hyperparameter Modeling**: Hierarchical structure for population-level parameters

## Workflow Overview

The typical BMLM workflow consists of the following steps:

1. **Configuration**: Set up your project using the reporting framework and configure BMLM-specific settings
2. **Model Setup**: Define parameters, priors, and model error specifications
3. **Optimization**: Initialize and run the optimization process
4. **Monitoring**: Check optimization progress using diagnostic plots
5. **Export Results**: Transfer optimized parameters back to the reporting framework

### Workflow and Data Flow Diagram

The following diagram visualizes the complete BMLM workflow, showing the relationships between different files and the data flow through the optimization process:

```{r workflow-diagram, echo=FALSE, eval=TRUE, fig.width=20, fig.height=12}
library(DiagrammeR)
grViz("
digraph BMLM_Workflow {
  graph [rankdir=TB, fontname=Helvetica, compound=true]
  node [shape=box, style=filled, fontname=Helvetica, margin=0.2]
  edge [fontname=Helvetica]
  
  # Styling
  node [fillcolor='#E8F4F8']

  # Input files
  subgraph cluster_input {
    label='Input Files (User Provided)'
    style=filled
    fillcolor='#FFF4E6'

    snapshot [label='Snapshot\n(.json)', fillcolor='#FFE6E6']
    observed [label='Observed Data\n(from Reporting Framework)', fillcolor='#FFE6E6']
    rf_config [label='Reporting Framework\nScenarios', fillcolor='#FFE6E6']
  }

  # Configuration phase
  subgraph cluster_config {
    label='Configuration Phase'
    style=filled
    fillcolor='#E6F3E6'

    add_config [label='addBMLMPConfiguration()', shape=ellipse, fillcolor='#C8E6C9']
    bmlm_xlsx [label='BMLMConfiguration.xlsx\n(Parameter definitions,\nMapped paths,\nModel errors)', fillcolor='#A5D6A7']
    config_priors [label='configurePriors()', shape=ellipse, fillcolor='#C8E6C9']
  }

  # Optimization initialization
  subgraph cluster_init {
    label='Optimization Initialization'
    style=filled
    fillcolor='#E1F5FE'

    bmlm_new [label='BMLMOptimization$new()', shape=ellipse, fillcolor='#B3E5FC']
    output_dir [label='Output Directory\nBMLMOptimization/<runName>/', fillcolor='#81D4FA']

    # Configuration files in output
    data_csv [label='data.csv', fillcolor='#4FC3F7']
    prior_csv [label='prior.csv', fillcolor='#4FC3F7']
    start_csv [label='startValues.csv', fillcolor='#4FC3F7']
    mapped_csv [label='mappedPaths.csv', fillcolor='#4FC3F7']
  }

  # Optimization execution
  subgraph cluster_optim {
    label='Optimization Execution'
    style=filled
    fillcolor='#F3E5F5'

    start_optim [label='startOptimization()', shape=ellipse, fillcolor='#E1BEE7']

    # Intermediate files
    conv_csv [label='convergence.csv\n(iteration history)', fillcolor='#CE93D8']
    status_rds [label='status.RDS\n(running status)', fillcolor='#CE93D8']
    optim_rds [label='optimStatus.RDS\n(current params)', fillcolor='#CE93D8']
    best_rds [label='bestOptimStatus.RDS\n(best params)', fillcolor='#CE93D8']
    pred_rds [label='bestPrediction.RDS\n(predictions)', fillcolor='#CE93D8']
    log_txt [label='optimization_log.txt\n(detailed log)', fillcolor='#CE93D8']
  }

  # Monitoring
  subgraph cluster_monitor {
    label='Progress Monitoring'
    style=filled
    fillcolor='#FFF9C4'

    check_conv [label='check\nConvergence()', shape=ellipse, fillcolor='#FFF59D']
    check_pred [label='check\nPredictedVsObserved()', shape=ellipse, fillcolor='#FFF59D']
    check_res [label='check\nResiduals()', shape=ellipse, fillcolor='#FFF59D']
    check_dist [label='check\nDistributions()', shape=ellipse, fillcolor='#FFF59D']
  }

  # Export phase
  subgraph cluster_export {
    label='Export Results'
    style=filled
    fillcolor='#E0F2F1'

    export_pop [label='export\nResultAsPopulation()', shape=ellipse, fillcolor='#B2DFDB']
    export_params [label='export\nModelParametersToConfigTables()', shape=ellipse, fillcolor='#B2DFDB']
    export_ind [label='export\nIndividualValuesToConfigTable()', shape=ellipse, fillcolor='#B2DFDB']
    export_pkml [label='export\nIndividualResultsToPkml()', shape=ellipse, fillcolor='#B2DFDB']
  }

  # Output files back to reporting framework
  subgraph cluster_output {
    label='Output Files (Back to Reporting Framework)'
    style=filled
    fillcolor='#FFE0B2'

    pop_files [label='Population CSV files\n(optimized individuals)', fillcolor='#FFCC80']
    model_params [label='ModelParameters.xlsx\n(global params, medians)', fillcolor='#FFCC80']
    pop_xlsx [label='Populations.xlsx\n(distribution params)', fillcolor='#FFCC80']
    ind_table [label='Individuals table\n(updated values)', fillcolor='#FFCC80']
    pkml_files [label='PKML files\n(individual models)', fillcolor='#FFCC80']
  }

  # Main workflow edges
  snapshot -> add_config
  add_config -> bmlm_xlsx

  bmlm_xlsx -> config_priors
  observed -> config_priors
  config_priors -> bmlm_xlsx [label='Updates\nPriors &\nStartValues', style=dashed]

  bmlm_xlsx -> bmlm_new
  observed -> bmlm_new
  rf_config -> bmlm_new

  bmlm_new -> output_dir
  output_dir -> data_csv
  output_dir -> prior_csv
  output_dir -> start_csv
  output_dir -> mapped_csv

  data_csv -> start_optim
  prior_csv -> start_optim
  start_csv -> start_optim
  mapped_csv -> start_optim

  start_optim -> conv_csv
  start_optim -> status_rds
  start_optim -> optim_rds
  start_optim -> best_rds
  start_optim -> pred_rds
  start_optim -> log_txt

  # Monitoring reads from intermediate files
  conv_csv -> check_conv [style=dashed]
  best_rds -> check_pred [style=dashed]
  pred_rds -> check_pred [style=dashed]
  best_rds -> check_res [style=dashed]
  pred_rds -> check_res [style=dashed]
  best_rds -> check_dist [style=dashed]

  # Export phase
  best_rds -> export_pop
  best_rds -> export_params
  best_rds -> export_ind
  best_rds -> export_pkml
  
  export_pop -> pop_files
  export_params -> model_params
  export_params -> pop_xlsx
  export_ind -> ind_table
  export_pkml -> pkml_files

  # Legend
  subgraph cluster_legend {
    label='Legend'
    style=filled
    fillcolor='white'

    leg_input [label='Input Files', fillcolor='#FFE6E6']
    leg_config [label='Configuration', fillcolor='#A5D6A7']
    leg_func [label='Functions', fillcolor='#C8E6C9', shape=ellipse]
    leg_temp [label='Intermediate Files', fillcolor='#CE93D8']
    leg_output [label='Output Files', fillcolor='#FFCC80']
  }

  {rank=same; leg_input; leg_config; leg_func; leg_temp; leg_output}
}
", width=1600, height=1000)
```

**Key File Dependencies:**

**Configuration Files (Created During Setup):**
- `BMLMConfiguration.xlsx`: Central configuration with multiple sheets (ParameterDefinition, ParameterMappedPaths, ModelError, Priors, IndividualStartValues)
- `data.csv`, `prior.csv`, `startValues.csv`, `mappedPaths.csv`: Extracted from configuration for optimization

**Intermediate Files (Updated During Optimization):**
- `convergence.csv`: Tracks log-likelihood and parameters at each iteration
- `status.RDS`: Current status (initialized/running/stopped)
- `optimStatus.RDS`: Current parameter values
- `bestOptimStatus.RDS`: Best parameters found so far
- `bestPrediction.RDS`: Model predictions with best parameters
- `optimization_log.txt`: Detailed execution log

**Final Output (After Optimization):**
- `result.RDS`: Complete optimization results
- Population CSV files: Optimized individual parameter values
- `ModelParameters.xlsx`: Global parameters and hyperparameter medians
- `Populations.xlsx`: Distribution parameters for population variability
- PKML files: Individual-specific models for PK-Sim/MoBi

## Integration with {ospsuite.reportingframework}

Before using BMLM, you need to set up your project configuration using the {ospsuite.reportingframework} package. This package allows you to define:

- Data sources
- Model specifications
- Simulation scenarios
- Observed data for parameter estimation

The BMLM package extends this framework by adding parameter identification capabilities through Bayesian multilevel modeling.

## Using the BMLM Add-In

The {ospsuite.bmlm} package includes an RStudio add-in called **Insert BMLM Template** that allows you to easily insert a template for a Bayesian Multilevel Model workflow into your existing reporting framework workflow. This feature saves time and ensures that your code follows a consistent structure.

### Accessing the Add-In

To use the add-in:

1. In RStudio, go to the **Addins** menu in the toolbar
2. Select **Insert BMLM Template** from the list
3. The template code will be inserted at your cursor position

### Template Workflow

The template includes:

- Configuration setup with snapshot loading
- Prior configuration
- Optimization initialization and execution
- Progress checking with diagnostic plots
- Results export to various formats

## Available Vignettes

This package provides several vignettes to help you get started and make the most of the BMLM functionality:

### Configuration
- **BMLM Configuration** ([vignette](02-configuration.html)): Detailed guidance on setting up the BMLM configuration file, including parameter definitions, mappings, model errors, and priors

### Optimization and Analysis
- **Running Optimization** ([vignette](03-optimization.html)): How to initialize the optimization object, start optimization runs, monitor progress with plotting functions, and export results back to the reporting framework

### Theoretical Background
- **Likelihood Components and Error Models** ([vignette](04-likelihood-components.html)): Mathematical formulation of the objective function, likelihood components, error models, and maximum likelihood estimation

## Quick Start Example

Here's a minimal example to get you started (assumes you have already set up a reporting framework project):

```{r, eval=FALSE}
# The workflow assumes the project configuration was initialized with the reporting engine
# and the following variables are defined
# - projectConfiguration
# - scenarioList
# - dataObserved

# Add BMLM configuration
projectConfiguration <- addBMLMPConfiguration(
  projectConfiguration,
  snapshotFile = file.path(projectConfiguration$modelFolder, "mySnapshot.json"),
  nameOfParameterIdentification = "myParameterIdentification"
)

# Configure priors
configurePriors(
  projectConfiguration = projectConfiguration,
  dataObserved = dataObserved
)

# Initialize optimization
myRun <- BMLMOptimization$new(
  projectConfiguration = projectConfiguration,
  runName = "myRun",
  scenarioList = scenarioList,
  dataObserved = dataObserved
)

# Evaluate initial values
myRun$evaluateInitialValues()

# Start optimization
myRun$startOptimization(
  method = "BFGS",
  control = list(maxit = 1000)
)

# Check progress
myRun$checkConvergence()
myRun$checkPredictedVsObserved()

# Export results
myRun$exportResultAsPopulation(projectConfiguration)
```

## Getting Help

For more detailed information on specific topics:

- **Configuration questions**: See the [Configuration vignette](02-configuration.html)
- **Optimization issues**: See the [Optimization vignette](03-optimization.html)
- **Mathematical background**: See the [Likelihood Components vignette](04-likelihood-components.html)
- **Function documentation**: Use `?BMLMOptimization` or `?configurePriors` in R
- **Package issues**: Visit the [GitHub repository](https://github.com/Open-Systems-Pharmacology/OSPSuite.BMLM)

## Next Steps

Now that you understand the basics of {ospsuite.bmlm}, we recommend:

1. Reading the [Configuration vignette](02-configuration.html) to learn how to set up your BMLM project
2. Following the [Optimization vignette](03-optimization.html) to run your first optimization
3. Consulting the [Likelihood Components vignette](04-likelihood-components.html) to understand the mathematical foundations

## Conclusion

The {ospsuite.bmlm} package provides a streamlined approach to fitting Bayesian multilevel models within the OSPSuite framework. By integrating with the reporting framework and providing flexible configuration options, it enables efficient parameter identification for pharmacometric models. We encourage you to explore the additional vignettes and consider contributing to the development of the package to enhance its capabilities.
