---
title: "BMLM Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BMLM Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(ospsuite.bmlm)
```

## Introduction

The `ospsuite.bmlm` package is designed to facilitate the fitting of Bayesian multilevel models within the OSPSuite framework. One of the critical components of this package is the configuration file, `BMLMConfiguration.xlsx`, which serves as the foundation for defining the parameters, priors, and mappings necessary for model optimization and analysis.

The configuration file is structured into multiple sheets, each dedicated to specific aspects of the modeling process. By utilizing this file, users can efficiently set up their modeling environment, ensuring that all relevant parameters are correctly defined. This structured approach allows for flexibility in model configuration, enabling users to adapt the model to their specific data and research questions.

In this vignette, we will guide you through the workflow for configuring your BMLM project, detailing each step necessary to prepare the `BMLMConfiguration.xlsx` file. We will also provide a comprehensive description of the contents of this configuration file, including the purpose of each sheet and the parameters that can be defined within them.

## Configuration Workflow

Configuring your Bayesian Multilevel Model (BMLM) involves several key steps to ensure that all parameters and settings are correctly defined. Below is the recommended workflow for setting up your BMLM configuration:

### Step 1: Add the BMLM Configuration File

Start by adding the BMLM configuration file to your project configuration. You can load a snapshot file that contains predefined parameters or, if `snapshotFile = NULL`, add the empty template.

```{r, eval=FALSE, echo=TRUE}
# Add BMLM configuration file and set up parameter identification from snapshot
projectConfiguration <- addBMLMPConfiguration(
  projectConfiguration,
  snapshotFile = file.path(projectConfiguration$modelFolder, "mySnapshot.json"),
  nameOfParameterIdentification = "myParameterIdentification",
  overwrite = FALSE
)
```

This function creates the `BMLMConfiguration.xlsx` file in your project's configuration directory. If a snapshot file is provided, it will pre-populate the configuration with parameters from the snapshot.

### Step 2: Fill or Adjust the Configuration Sheets

Manually fill or adjust the following sheets in the `BMLMConfiguration.xlsx` file:

- **ParameterDefinition**: Define the parameters used in your model
- **ParameterMappedPaths**: Specify how model paths map to the defined parameters
- **ModelError**: Set up the model error parameters associated with the outputs

These sheets form the basis of your BMLM configuration and must be completed before proceeding.

### Step 3: Configure Priors

After the sheets are populated, call the `configurePriors` function to automatically fill in the **Priors** and **IndividualStartValues** sheets based on your BMLM configuration.

```{r, eval=FALSE, echo=TRUE}
# Configure Priors based on BMLM configuration file
configurePriors(
  projectConfiguration = projectConfiguration,
  dataObserved = dataObserved
)
```

This function analyzes your parameter definitions and observed data to generate appropriate prior distributions and starting values for individual parameters.

### Step 4: Adjust Priors and Start Values

Finally, review and manually adjust the **Priors** and **IndividualStartValues** sheets as necessary to ensure that they reflect the desired prior distributions and starting values for your parameters. This step allows you to incorporate domain knowledge and previous experimental results into your model.

## Configuration File Structure

The `BMLMConfiguration.xlsx` file contains multiple sheets, each serving a specific purpose. Below, we describe each sheet in detail.

## ParameterDefinition Sheet

The **ParameterDefinition** sheet is the primary sheet for configuring the parameters used in the Bayesian multilevel model. It serves as the basis for the `Prior` and `StartValue` generation. After these sheets are generated, this sheet is not actively used during optimization but serves as documentation.

### Columns

| Column Name             | Description                                                                                                          |
|-------------------------|----------------------------------------------------------------------------------------------------------------------|
| **Name**                | The unique identifier and display name of the parameter in reports. Must be unique.                                 |
| **ValueMode**           | Specifies if the parameter is "global" (same value for all individuals) or "individual" (varies by individual).    |
| **Distribution**        | The distribution for parameters with ValueMode "individual". Can be selected from various statistical distributions (e.g., normal, lognormal, weibull). |
| **CategoricCovariate**  | Indicates a categorical covariate (e.g., "SEX"). If not empty, a distinct distribution is fitted for each subgroup defined by the covariate. Only applicable for parameters with ValueMode "individual". |
| **Unit**                | The unit of measurement for the parameter.                                                                          |
| **StartValue**          | The initial value for the parameter used to begin optimization.                                                    |
| **MinValue**            | The minimum allowable value for the parameter (mandatory).                                                         |
| **MaxValue**            | The maximum allowable value for the parameter (mandatory).                                                         |
| **Scaling**             | Indicates if the parameter should be log-transformed during optimization (set to "log") or kept linear (set to "linear"). |
| **UseAsFactor**         | If TRUE (1), the fitted value is applied as a factor to all mapped parameters. If FALSE (0), the fitted value replaces the parameter value. |
| **Comment**             | Additional comments or notes regarding the parameter.                                                              |

### Example Configuration

Here is an example of how a configuration might look in the **ParameterDefinition** sheet:

| Name           | ValueMode  | Distribution | CategoricCovariate | Unit  | StartValue | MinValue | MaxValue | Scaling | UseAsFactor | Comment                         |
|----------------|------------|--------------|---------------------|-------|------------|----------|----------|---------|-------------|---------------------------------|
| Lipophilicity  | global     |              |                     |       | 3.5        | 0        | 10       | linear  | 0           | Compound lipophilicity          |
| CL             | individual | lnorm_geomean| SEX                 | L/h   | 10         | 1        | 100      | log     | 1           | Clearance by sex group          |
| GET            | individual | weibull      |                     | min   | 30         | 5        | 120      | linear  | 0           | Gastric emptying time           |

### Notes

- **ValueMode**: Choose "global" for compound properties that don't vary between individuals. Choose "individual" for physiological parameters that vary in the population.
- **Distribution**: Only required for individual parameters. Common choices include normal, lnorm_geomean (log-normal), and weibull distributions.
- **CategoricCovariate**: Use this to model parameter distributions that differ by subgroups (e.g., males vs. females, different age groups).
- **Scaling**: Log scaling is often appropriate for parameters that span multiple orders of magnitude.
- **UseAsFactor**: Set to 1 when you want to scale existing parameter values rather than replace them completely.

## ParameterMappedPaths Sheet

The **ParameterMappedPaths** sheet defines the mapping of configuration parameters to the corresponding model parameters. This ensures that the optimization correctly updates the model parameters.

### Columns

| Column Name          | Description                                                                                                          |
|----------------------|----------------------------------------------------------------------------------------------------------------------|
| **Name**             | The identifier of the parameter (must match a Name from ParameterDefinition).                                       |
| **LinkedParameters** | The model path(s) of all parameters that this configuration parameter maps to. Use pipe notation (e.g., "Organism\|Weight"). |
| **Scenarios**        | The scenario(s) in which these parameters are used. Can be a comma-separated list. If empty, applies to all scenarios that include this parameter. |
| **Comment**          | Additional comments regarding the parameter mappings.                                                                |

### Example Configuration

| Name          | LinkedParameters                                    | Scenarios | Comment                              |
|---------------|-----------------------------------------------------|-----------|--------------------------------------|
| GET           | Organism\|Lumen\|Stomach\|Gastric emptying time     |           | Mapping for gastric emptying time    |
| Lipophilicity | MyCompound\|Lipophilicity                           |           | Compound lipophilicity parameter     |
| CL            | MyCompound\|Clearance                               | Scenario1 | Clearance specific to Scenario1      |

### Notes

- Use the exact path syntax as it appears in the OSPSuite model (with `|` separators).
- You can map a single configuration parameter to multiple model parameters if needed.
- If a parameter appears in multiple scenarios but with different mappings, create separate rows with different Scenarios values.

## ModelError Sheet

The **ModelError** sheet specifies the model error parameters associated with the outputs from your model. This is important for quantifying uncertainty in the model predictions.

### Columns

| Column Name       | Description                                                                                                          |
|-------------------|----------------------------------------------------------------------------------------------------------------------|
| **OutputPathId**  | The identifier for the output path as defined in the reporting framework.                                          |
| **ErrorModel**    | The error model type: "proportional", "absolute", or "log_absolute".                                               |
| **ModelErrorId**  | Identifier for the error parameter. Outputs with the same ModelErrorId share the same sigma (error magnitude) parameter. |
| **Comment**       | Additional comments regarding the model error parameters.                                                           |

### Example Configuration

| OutputPathId | ErrorModel   | ModelErrorId  | Comment                          |
|--------------|--------------|---------------|----------------------------------|
| Plasma_PK    | log_absolute | sigma_plasma  | Log-scale error for plasma PK    |
| Urine_PK     | absolute     | sigma_urine   | Absolute error for urine data    |
| Metabolite   | proportional | sigma_met     | Proportional error for metabolite|

### Error Model Types

- **absolute**: Error is constant across all concentration ranges. Appropriate when measurement error is constant.
- **proportional**: Error increases proportionally with the measured value. Common for concentration measurements.
- **log_absolute**: Error is constant on a log scale. Appropriate when relative error is constant.

### Notes

- Multiple outputs can share the same ModelErrorId to share a single error parameter.
- The choice of error model should reflect the characteristics of your measurement method.

## Priors Sheet

The **Priors** sheet defines the prior distributions for parameters. This sheet is automatically generated by `configurePriors()` but can be manually adjusted.

### Columns

| Column Name                          | Description                                                                                                          |
|--------------------------------------|----------------------------------------------------------------------------------------------------------------------|
| **Name**                             | The name of the parameter.                                                                                           |
| **ValueMode**                        | Specifies if the prior is "global", "hyperparameter", or "outputError".                                             |
| **HyperParameter**                   | For hyperparameters: indicates which distribution parameter this row defines (e.g., "geomean", "geosd", "mean", "sd"). |
| **CategoricCovariate**               | Categorical covariate for which this hyperparameter applies (if applicable).                                         |
| **StartValue**                       | Initial value for the parameter.                                                                                     |
| **MinValue**                         | Minimum limit for the parameter.                                                                                     |
| **MaxValue**                         | Maximum limit for the parameter.                                                                                     |
| **Distribution of Individual Values**| The distribution type for individual parameters (e.g., "lnorm_geomean", "normal").                                  |
| **UseAsFactor**                      | If TRUE (1), the fitted value is applied as a factor.                                                               |
| **Unit**                             | The unit of measurement.                                                                                             |
| **Scaling**                          | Parameter scaling ("log" or "linear").                                                                               |
| **Prior Distribution**               | The prior distribution type (e.g., "flat", "norm", "lnorm_geomean").                                               |
| **P1_type**, **P1_value**            | First parameter of the prior distribution.                                                                           |
| **P2_type**, **P2_value**            | Second parameter of the prior distribution (if applicable).                                                         |
| **P3_type**, **P3_value**            | Third parameter of the prior distribution (if applicable).                                                          |
| **Comment**                          | Additional comments.                                                                                                 |

### Example Configuration

| Name | ValueMode      | HyperParameter | CategoricCovariate | StartValue | MinValue | MaxValue | Distribution of Individual Values | UseAsFactor | Unit | Scaling | Prior Distribution | P1_type | P1_value | P2_type | P2_value | Comment              |
|------|----------------|----------------|---------------------|------------|----------|----------|-----------------------------------|-------------|------|---------|-------------------|---------|----------|---------|----------|----------------------|
| Lipophilicity | global    |                |                     | 3.5        | 0        | 10       |                                   | 0           |      | linear  | flat              |         |          |         |          | Flat prior           |
| CL   | hyperparameter | geomean        | SEX                 | 10         | 1        | 100      | lnorm_geomean                     | 1           | L/h  | log     | lnorm_geomean     | geomean | 10       | geosd   | 1.5      | Prior on CL mean     |
| CL   | hyperparameter | geosd          | SEX                 | 1.5        | 1        | 5        | lnorm_geomean                     | 1           | L/h  | linear  | flat              |         |          |         |          | Flat prior on CV     |
| sigma_plasma | outputError |              |                     | 0.3        | 0.01     | 2        |                                   | 0           |      | linear  | flat              |         |          |         |          | Error parameter      |

### Notes

- **flat prior**: No prior information; use `flat` as the Prior Distribution.
- **Hyperparameters**: For individual parameters, you need separate rows for each distribution parameter (e.g., mean and standard deviation).
- **Prior distributions**: Can use any distribution available in R's `stats` package (dnorm, dlnorm, etc.) or the custom `lnorm_geomean` distribution.

## IndividualStartValues Sheet

The **IndividualStartValues** sheet specifies the initial values for individual parameters. This sheet is automatically generated by `configurePriors()` based on the observed data but can be manually adjusted.

### Columns

| Column Name              | Description                                                                                                          |
|--------------------------|----------------------------------------------------------------------------------------------------------------------|
| **Name**                 | The name of the parameter.                                                                                           |
| **IndividualId**         | The identifier for the individual (must match IDs in observed data).                                                |
| **StartValue**           | The initial value for this individual's parameter.                                                                  |
| **CategoricCovariate**   | The value of the categorical covariate for this individual (if applicable).                                         |
| **MinValue**             | The minimum allowable value for the parameter (mandatory).                                                          |
| **MaxValue**             | The maximum allowable value for the parameter (mandatory).                                                          |
| **Scaling**              | Parameter scaling ("log" or "linear").                                                                               |
| **Comment**              | Additional comments.                                                                                                 |

### Example Configuration

| Name | IndividualId | StartValue | CategoricCovariate | MinValue | MaxValue | Scaling | Comment                              |
|------|--------------|------------|--------------------|----------|----------|---------|--------------------------------------|
| CL   | Individual_01| 12         | Male               | 1        | 100      | log     | Starting value for Individual 01     |
| CL   | Individual_02| 8          | Female             | 1        | 100      | log     | Starting value for Individual 02     |
| GET  | Individual_01| 30         |                    | 5        | 120      | linear  | GET for Individual 01                |
| GET  | Individual_02| 45         |                    | 5        | 120      | linear  | GET for Individual 02                |

### Notes

- Each individual must have a start value for each individual parameter defined in ParameterDefinition.
- Start values should be physiologically reasonable to aid optimization convergence.
- The `configurePriors()` function can generate reasonable starting values automatically, but manual refinement may improve optimization performance.

## Best Practices

### General Configuration Tips

1. **Start Simple**: Begin with a minimal set of parameters and add complexity gradually.
2. **Check Units**: Ensure all units are consistent across configuration sheets.
3. **Use Informative Priors**: When available, incorporate prior knowledge from literature or previous studies.
4. **Validate Mappings**: Double-check that ParameterMappedPaths correctly references model parameters.

### Parameter Definition

1. **Choose Appropriate Distributions**: Select distributions that match the biological/physical nature of the parameter.
2. **Set Reasonable Bounds**: MinValue and MaxValue should encompass physiologically plausible ranges but not be unnecessarily wide.
3. **Consider Scaling**: Use log scaling for parameters that span orders of magnitude.

### Prior Selection

1. **Flat Priors**: Use when no prior information is available.
2. **Informative Priors**: Use normal or log-normal priors when you have prior knowledge.
3. **Weakly Informative Priors**: Use when you have some information but want to let the data dominate.

### Starting Values

1. **Physiologically Reasonable**: Choose values that are biologically plausible.
2. **Avoid Boundaries**: Don't start exactly at min or max values.
3. **Use Literature Values**: When available, use published parameter values as starting points.

## Common Issues and Solutions

### Issue: Parameters not updating during optimization

**Solution**: Check that:
- ParameterMappedPaths correctly references the model parameter paths
- Scenarios column includes the relevant scenarios
- Parameter names match exactly between sheets

### Issue: Optimization fails to converge

**Solution**:
- Review starting values for reasonableness
- Check that MinValue and MaxValue bounds are appropriate
- Consider changing parameter scaling (log vs. linear)
- Try different optimization methods

### Issue: Error parameter values are unrealistic

**Solution**:
- Review the error model choice (proportional vs. absolute vs. log_absolute)
- Adjust the bounds for error parameters in the Priors sheet
- Check that data quality is sufficient for the chosen error model

## Next Steps

After completing your configuration:

1. **Validate Configuration**: Use `evaluateInitialValues()` to check that your model runs with the configured parameters.
2. **Start Optimization**: Proceed to the [Optimization vignette](03-optimization.html) to learn how to run the optimization.
3. **Understand Mathematical Details**: Consult the [Mathematical Details vignette](04-mathematical-details.html) for details on the mathematical foundations.

## Conclusion

By following the outlined workflow and understanding the structure of the `BMLMConfiguration.xlsx` file, you can effectively configure your Bayesian Multilevel Models for optimal performance. The configuration process may seem complex at first, but the structured approach ensures that all necessary information is captured systematically. For further details on running optimization and monitoring progress, please refer to the [Optimization vignette](03-optimization.html).
