---
title: "Theory and Methodology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory and Methodology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rmarkdown.html_vignette.check_title = FALSE
)
```

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(ospsuite.bmlm)
```

## Introduction

This vignette provides a comprehensive overview of the theoretical foundations underlying the Bayesian multilevel modeling approach implemented in the `ospsuite.bmlm` package. We explain the mathematical formulation of the objective function, the optimization algorithm, and the concepts of internal optimization and parameter scaling.

Understanding these theoretical concepts will help you:
- Make informed decisions about model configuration
- Interpret optimization results correctly
- Troubleshoot optimization problems
- Choose appropriate priors and error models

## Bayesian Multilevel Models

Bayesian multilevel models, also known as hierarchical models, are designed for parameter identification using individual study data. They are particularly useful when:

- Data come from multiple individuals or groups
- Parameters vary between individuals (random effects)
- Some parameters are shared across individuals (fixed effects)
- You want to borrow strength across individuals for better parameter estimates

### Model Structure

The hierarchical structure consists of multiple levels:

1. **Individual Level**: Each individual has their own parameter values
2. **Population Level**: Distribution parameters (hyperparameters) describe variability across individuals  
3. **Global Level**: Parameters that are the same for all individuals (e.g., compound properties)

This structure allows the model to:
- Estimate individual-specific parameters
- Characterize population variability
- Identify parameters that are constant across the population

## Maximum Likelihood Estimation

Maximum likelihood estimation (MLE) is the core statistical method used for parameter identification in this package. MLE seeks to find parameter values that make the observed data most probable.

### Objective Function

The goal of MLE is to find the parameter values that maximize the likelihood function:

$$ \hat{\theta} = \underset{\theta \in \Theta}{\operatorname{arg\,max}} \mathcal{L}(\theta, \mathbf{y}, \mathbf{p})$$

Where:
- $\theta$: Vector of all adjustable parameters
- $\mathbf{y}$: Vector of observed data across all individuals and outputs
- $\mathbf{p}$: Prior distributions on parameters

### Parameter Vector $\theta$

The parameter vector $\theta$ contains:

1. **Global Parameters**: One scalar per global parameter, denoted $\theta_k$
2. **Individual Parameters**: For parameters with hyperparameters:
   - Individual values $\theta_{k_i}$ for each individual $i$
   - Distribution parameters (hyperparameters) $\theta_{k_p}$
3. **Model Error Parameters**: One or more error parameters $\epsilon_j$ for each output type

### Conversion to Minimization

For use in optimization algorithms, the maximization problem is converted to minimization:

$$ \hat{\theta} = \underset{\theta \in \Theta}{\operatorname{arg\,min}} \left(-\log\left(\mathcal{L}(\theta, \mathbf{y}, \mathbf{p})\right)\right)$$

This negative log-likelihood is the objective function that the optimization algorithm seeks to minimize.

## Construction of the Log-Likelihood

The total log-likelihood is constructed as the sum of three components:

$$ \log(\mathcal{L}(\theta, \mathbf{y}, \mathbf{p})) = \log(\mathcal{L}_t(\theta, \mathbf{y})) + \log(\mathcal{L}_h(\theta)) + \log(\mathcal{L}_p(\theta, \mathbf{p}))$$

Where:
- $\mathcal{L}_t$: Likelihood of observed data given simulated time profiles
- $\mathcal{L}_h$: Likelihood of hyperparameters (for individual parameters)
- $\mathcal{L}_p$: Likelihood of parameter estimates given prior distributions

Let's examine each component in detail.

## Likelihood for Time Profiles $\log(\mathcal{L}_t(\theta, \mathbf{y}))$

The likelihood for observed data is calculated as:

$$ \log(\mathcal{L}_t(\theta, \mathbf{y})) = \sum_{i,j}\left(\log(\mathcal{L}u_{i,j}(\theta, \mathbf{y})) + \log(\mathcal{L}c_{i,j}(\theta, \mathbf{y}))\right)$$

Where:
- $i$: Individual index
- $j$: Observer/output index (e.g., plasma concentration, metabolite, urine data)
- $\mathcal{L}u(\theta, \mathbf{y})$: Likelihood for uncensored data points (values above LLOQ)
- $\mathcal{L}c(\theta, \mathbf{y})$: Likelihood for censored data points (values below LLOQ)

The package supports three error models, each with different formulations for the likelihood.

### Additive Error Model

In the additive error model, the measurement error is constant across all concentration levels.

#### Uncensored Data Points

$$\log(\mathcal{L}u_{i,j}(\theta, \mathbf{y})) = -\frac{1}{2}\sum_{t_u}\left(\frac{(\mathbf{y}_{i,j}(t) - f_{i,j}(\theta,t))^2}{\epsilon_j^2} + \log(2\pi \epsilon_j^2)\right) - \log\left(1-\Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j}\right)\right)$$

#### Censored Data Points

$$\log(\mathcal{L}c_{i,j}(\theta, \mathbf{y})) = \sum_{t_c}\left(\log\left(\Phi\left(\frac{lloq_{j}(t) - f_{i,j}(\theta,t)}{\epsilon_j}\right) - \Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j}\right)\right) - \log\left(1-\Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j}\right)\right)\right)$$

**Notation:**
- $\mathbf{y}_{i,j}(t)$: Observed data point for individual $i$ and output $j$ at time $t$
- $f_{i,j}(\theta,t)$: Simulated (predicted) data point
- $lloq_{j}(t)$: Lower limit of quantification
- $LB$: Physiological lower bound (typically 0)
- $\epsilon_j$: Model error (standard deviation) for output $j$
- $\Phi(x)$: Cumulative distribution function of the standard normal distribution

**When to Use:** When measurement error is approximately constant regardless of concentration level.

### Proportional Error Model

In the proportional error model, the measurement error increases proportionally with the measured value.

#### Uncensored Data Points

$$\log(\mathcal{L}u_{i,j}(\theta, \mathbf{y})) = -\frac{1}{2}\sum_{t_u}\left(\frac{(\mathbf{y}_{i,j}(t) - f_{i,j}(\theta,t))^2}{\epsilon_j^2 f_{i,j}(\theta,t)^2} + \log(2\pi \epsilon_j^2 f_{i,j}(\theta,t)^2)\right) - \log\left(1-\Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j f_{i,j}(\theta,t)}\right)\right)$$

#### Censored Data Points

$$\log(\mathcal{L}c_{i,j}(\theta, \mathbf{y})) = \sum_{t_c}\left(\log\left(\Phi\left(\frac{lloq_{j}(t) - f_{i,j}(\theta,t)}{\epsilon_j f_{i,j}(\theta,t)}\right) - \Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j f_{i,j}(\theta,t)}\right)\right) - \log\left(1-\Phi\left(\frac{LB - f_{i,j}(\theta,t)}{\epsilon_j f_{i,j}(\theta,t)}\right)\right)\right)$$

**When to Use:** When measurement error is proportional to concentration (constant coefficient of variation).

### Logarithmic Error Model

In the logarithmic error model, the error is constant on a log scale, which is equivalent to a constant relative error.

#### Uncensored Data Points

$$\log(\mathcal{L}u_{i,j}(\theta, \mathbf{y})) = -\frac{1}{2}\sum_{t_u}\left(\frac{(\log(\mathbf{y}_{i,j}(t)) - \log(f_{i,j}(\theta,t)))^2}{\epsilon_j^2} + \log(2\pi \epsilon_j^2 \mathbf{y}_{i,j}(t)^2)\right) - \log\left(1-\Phi\left(\frac{\log(LB) - \log(f_{i,j}(\theta,t))}{\epsilon_j}\right)\right)$$

#### Censored Data Points

$$\log(\mathcal{L}c_{i,j}(\theta, \mathbf{y})) = \sum_{t_c}\left(\log\left(\Phi\left(\frac{\log(lloq_{j}(t)) - \log(f_{i,j}(\theta,t))}{\epsilon_j}\right) - \Phi\left(\frac{\log(LB) - \log(f_{i,j}(\theta,t))}{\epsilon_j}\right)\right) - \log\left(1-\Phi\left(\frac{\log(LB) - \log(f_{i,j}(\theta,t))}{\epsilon_j}\right)\right)\right)$$

**When to Use:** When error is approximately constant on a logarithmic scale; often appropriate for concentration data spanning multiple orders of magnitude.

### Cumulative Normal Distribution

The cumulative distribution function of the standard normal distribution is defined as:

$$\Phi(x) = \frac{1}{\sqrt{2\pi}}\int_{-\infty}^x e^{-\frac{t^2}{2}} dt$$

This function is used to handle:
- Lower limits of quantification (censored data)
- Physiological lower bounds (e.g., concentrations cannot be negative)

## Likelihood for Hyperparameters $\log(\mathcal{L}_h(\theta))$

For individual parameters that are modeled with a population distribution, the hyperparameter likelihood quantifies how well the individual values conform to the assumed distribution.

$$\log(\mathcal{L}_h(\theta)) = \sum_k \log(\mathcal{L}_{h,k}(\theta)) = \sum_{k \in K_h} \log(\phi(\theta_{k}))$$

Where:
- $K_h$: Set of parameters with hyperparameters
- $\phi(\theta_k)$: Probability density function of the hyperparameter distribution
- $\theta_{k}$: Individual parameter values and distribution parameters for parameter $k$

### Example: Multimodal Normal Distribution

For a normal distribution with multiple modes (groups):

$$\log(\mathcal{L}_{h,k}(\theta)) = \sum_g\left(-\frac{1}{2\sigma_{k,g}^2}\sum_{i \in g}(value_{k,i} - \mu_{k,g})^2 - \frac{n_g}{2}\log(2\pi\sigma_{k,g}^2)\right)$$

Where:
- $g$: Group/mode index
- $i \in g$: Individuals belonging to group $g$
- $value_{k,i}$: Individual value for parameter $k$ and individual $i$
- $\mu_{k,g}$: Mean of hyperparameter for group $g$
- $\sigma_{k,g}$: Standard deviation of hyperparameter for group $g$
- $n_g$: Number of individuals in group $g$

This formulation allows for:
- Different parameter distributions for different subgroups (e.g., by sex, age group)
- Borrowing strength across individuals within each group
- Separate characterization of variability in each group

## Likelihood for Prior Distributions $\log(\mathcal{L}_p(\theta, \mathbf{p}))$

Prior distributions allow incorporation of previous knowledge about parameter values:

$$\log(\mathcal{L}_p(\theta, \mathbf{p})) = \sum_{k \in K_p} \log(\mathcal{L}_{p,k}(\theta, \mathbf{p})) = \sum_{k \in K_p} \log(\Pi(\theta_{k}, \mathbf{p}))$$

Where:
- $K_p$: Set of parameters with defined priors
- $\Pi(\theta_k, \mathbf{p})$: Probability density function of the prior distribution

### Available Distributions

The package supports all distributions available in R's `stats` package:
- Normal: `dnorm`
- Log-normal: `dlnorm`
- Gamma: `dgamma`
- Beta: `dbeta`
- Uniform: `dunif`
- And many others

Additionally, the package provides:
- **`lnorm_geomean`**: Log-normal distribution parameterized by geometric mean and geometric standard deviation (instead of log-mean and log-SD)
- **`flat`**: Uninformative (flat) prior where $\log(\mathcal{L}_{p,k}) = 0$

### Choosing Priors

**Flat Priors:**
- Use when no prior information is available
- Let the data completely determine the parameter
- May lead to convergence issues for weakly identified parameters

**Informative Priors:**
- Incorporate literature values or previous experimental results
- Help stabilize estimation for weakly identified parameters
- Can introduce bias if prior information is incorrect

**Weakly Informative Priors:**
- Provide gentle constraints without strong assumptions
- Good balance between incorporating knowledge and letting data speak
- Often a good default choice

**Caution:** The choice of prior distribution can significantly impact results, especially with limited data. Always check sensitivity to prior specifications.

## Optimization Algorithm

The package uses R's `optim()` function for parameter optimization, which implements several algorithms for function minimization.

### Algorithm Overview

The basic workflow is:

1. **Parameter Transformation**: Transform parameters from their natural scale to an unconstrained optimization space
2. **Objective Function Evaluation**:
   - Update parameters in simulation models
   - Run simulations for all scenarios
   - Calculate log-likelihood components
   - (Optionally) Perform internal optimization
   - Return negative log-likelihood
3. **Parameter Update**: Optimization algorithm updates parameters
4. **Iteration**: Repeat steps 2-3 until convergence

### Supported Optimization Methods

**BFGS (Default):**
- Quasi-Newton method
- Uses gradient approximations
- Generally efficient and robust
- Good for smooth objective functions

**Nelder-Mead:**
- Simplex-based method
- Derivative-free
- More robust to non-smooth objectives
- Can be slower than BFGS

**CG (Conjugate Gradients):**
- Useful for large parameter spaces
- Lower memory requirements than BFGS
- May be slower to converge

**L-BFGS-B:**
- Limited-memory BFGS with box constraints
- Can explicitly handle parameter bounds
- Good for large problems

**SANN (Simulated Annealing):**
- Global optimization method
- Better at escaping local minima
- Much slower than gradient-based methods

### Convergence Criteria

Optimization stops when one of the following is met:

1. **Maximum iterations reached** (`maxit` in control parameters)
2. **Relative tolerance satisfied**: Change in objective function falls below `reltol`
3. **Absolute tolerance satisfied**: Objective function value falls below `abstol`

## Internal Optimization

Internal optimization is a key feature that can significantly improve optimization efficiency for hierarchical models.

### Concept

Parameters are divided into two groups with different characteristics:

**Group 1: Global and Individual Parameters**
- Require full model simulations for evaluation
- Often have complex, multimodal objective functions
- Computationally expensive to evaluate
- May have multiple local minima

**Group 2: Hyperparameters and Error Parameters**
- Do not require model simulations (work with simulation output)
- Have relatively simple, well-behaved objective functions
- Computationally cheap to evaluate
- Usually unimodal and smooth

### How Internal Optimization Works

When `withInternalOptimization = TRUE`:

1. **Outer Optimization Loop**:
   - Optimize global parameters and individual parameter values
   - Each iteration requires full model simulations
   
2. **Inner Optimization Loop** (performed within each outer iteration):
   - Fix global and individual parameters
   - Optimize only hyperparameters and error parameters
   - Uses simulation output from outer loop
   - Quick to converge due to simple objective function

### Benefits

**Efficiency:**
- Internal optimization is much faster than full model evaluation
- Allows hyperparameters to quickly find optimal values given current individual parameters
- Reduces total number of expensive outer iterations needed

**Improved Convergence:**
- Separating parameter types helps navigate complex optimization landscape
- Prevents hyperparameters from slowing down global/individual parameter optimization
- Often leads to better final solutions

**Practical Impact:**
- Can reduce total optimization time by 50% or more
- Particularly beneficial for models with many individuals
- Essential for complex hierarchical models

### When to Use

**Recommended:**
- Models with hyperparameters (individual parameters with population distributions)
- Large numbers of individuals (> 10)
- Complex hierarchical structure
- When standard optimization struggles to converge

**May Not Be Necessary:**
- Models with only global parameters
- Very small datasets (< 5 individuals)
- Simple parameter structures

## Parameter Scaling

Proper parameter scaling is critical for optimization performance. The package implements a two-stage scaling approach.

### Stage 1: Normalization to [0, 1]

Parameters are first scaled to the range [0, 1] based on their defined bounds.

#### Linear Scaling

For parameters with `scaling = "linear"`:

$$\text{scaledValue} = \frac{\text{value} - \text{minValue}}{\text{maxValue} - \text{minValue}}$$

**Properties:**
- Preserves linear relationships
- Uniform sensitivity across parameter range
- Appropriate for parameters that vary linearly

#### Logarithmic Scaling

For parameters with `scaling = "log"`:

$$\text{scaledValue} = \frac{\log(\text{value}) - \log(\text{minValue})}{\log(\text{maxValue}) - \log(\text{minValue})}$$

**Properties:**
- Appropriate for parameters spanning multiple orders of magnitude
- Provides equal sensitivity to relative changes
- Common for pharmacokinetic parameters (clearance, volume, etc.)

### Stage 2: Boundary Handling

After normalization, the package applies one of two methods to handle parameter bounds during optimization.

#### Hard Bounds (`SCALINGMETHOD$hardBounds`)

**Mechanism:**
- Parameters strictly constrained to [0, 1] after normalization
- Values outside bounds are invalid and return NA

**Properties:**
- Enforces strict physical or biological constraints
- Prevents exploration beyond specified ranges
- May trap optimization in local minima near boundaries

**When to Use:**
- Strong physical constraints exist (e.g., fractions between 0 and 1)
- Parameters absolutely cannot exceed bounds
- When you have high confidence in bound specifications

#### LogSig Scaling (`SCALINGMETHOD$logsig`)

**Mechanism:**
- Applies logistic transformation: $\sigma(x) = \frac{1}{1 + e^{-x}}$
- Maps any real number to (0, 1)
- Allows optimization to explore beyond nominal bounds

**Properties:**
- Removes hard boundaries from optimization space
- Allows temporary excursions beyond specified ranges
- Can help escape local minima near boundaries
- Final values are transformed back to original scale

**When to Use (Recommended in most cases):**
- Complex optimization landscapes
- Uncertainty about exact bound values
- When optimization hits boundaries frequently
- Hierarchical models where flexibility helps convergence

### Impact on Optimization

**Good Scaling:**
- All parameters have similar sensitivity
- Optimization converges smoothly
- Gradient-based methods work efficiently

**Poor Scaling:**
- Some parameters dominate objective function
- Slow convergence
- Optimization may miss optimal values
- Gradient-based methods struggle

**Recommendations:**
1. Use log scaling for parameters spanning orders of magnitude
2. Use LogSig boundary method for most applications
3. Adjust bounds if parameters consistently hit boundaries
4. Check parameter limit plots to verify appropriate scaling

## Practical Implications

Understanding the theory helps in several practical situations:

### Choosing Error Models

- Review your measurement process
- Consider concentration range (absolute vs. proportional)
- Check residual plots to validate choice

### Setting Priors

- Use flat priors when no information exists
- Use informative priors from literature when available
- Consider sensitivity analysis for prior impact

### Optimization Strategy

- Use internal optimization for hierarchical models
- Choose appropriate scaling for each parameter
- Start with BFGS, try other methods if needed

### Interpreting Results

- Check all three likelihood components
- Verify hyperparameters make sense
- Examine residuals for model adequacy

## Computational Considerations

### Simulation Time

- Most computational time is spent in model simulations
- Number of simulations = (number of individuals) × (number of outputs)
- Internal optimization doesn't add simulation time

### Optimization Time

- Depends on number of parameters
- BFGS typically needs 10-100 iterations per parameter
- Internal optimization reduces outer iterations needed

### Memory Requirements

- Convergence history saved for all parameters
- Predictions saved for best parameters
- Usually not a limiting factor

## Conclusion

The theoretical framework underlying `ospsuite.bmlm` combines:
- Maximum likelihood estimation for parameter identification
- Hierarchical modeling for population analysis
- Flexible prior specification for incorporating knowledge
- Efficient optimization through internal optimization
- Robust parameter scaling for numerical stability

Understanding these concepts helps you:
- Configure models appropriately
- Choose optimization settings wisely
- Interpret results correctly
- Troubleshoot problems effectively

For practical application of these concepts, refer to:
- [Configuration vignette](02-configuration.html) for setting up models
- [Optimization vignette](03-optimization.html) for running analyses
- [Introduction vignette](01-introduction.html) for workflow overview

## References

For more information on the statistical and computational methods:

1. Bates, D. M., et al. (2015). Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48.

2. Gelman, A., et al. (2013). Bayesian Data Analysis, Third Edition. CRC Press.

3. Nocedal, J., & Wright, S. J. (2006). Numerical Optimization, Second Edition. Springer.

4. Pinheiro, J. C., & Bates, D. M. (2000). Mixed-Effects Models in S and S-PLUS. Springer.

These references provide deeper insights into Bayesian hierarchical modeling, maximum likelihood estimation, and numerical optimization algorithms.
