---
title: "Configuring Optimization in Bayesian Hierarchical Models: Scaling Methods and Internal Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{configuringOptimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  comment = "#>"
)
```

```{r setup}
library(ospsuite.bmlm)
library(ggplot2)
library(data.table)

options(rmarkdown.html_vignette.check_title = FALSE)
```

## Introduction

In Bayesian hierarchical modeling, optimizing parameters effectively is crucial for achieving accurate model fits. The `BMLMOptimization` class in the `ospsuite.bmlm` package encapsulates the functionality for performing Bayesian Maximum Likelihood Model (BMLM) optimization, leveraging the `optim` function at its core for parameter estimation. This vignette focuses on configuring the optimization process, the `evaluateInitialValues` and `startOptimization` methods, and the importance of scaling methods.

This vignette concentrates on the more complex aspects of the optimization process, specifically the scaling methods and internal optimization strategies, which enhance the performance and accuracy of the optimization tasks. Please refer also to the function help.


## Overview of BMLMOptimization

The `BMLMOptimization` class manages the configuration, output directories, logging, and optimization processes. It provides methods to evaluate initial values and to start the optimization process.

### Key Methods

#### 1. `evaluateInitialValues`

The `evaluateInitialValues` method evaluates initial values for a given project configuration by running simulations for specified scenarios. This method logs the progress and results to both the console and a designated log file.


- **Purpose**: This method is essential for assessing the initial parameter estimates before the optimization begins. It helps identify potential issues with the model setup or data.
- **Functionality**: It runs simulations based on the initial values and logs the resulting log-likelihoods. This provides a baseline for further optimization.

#### 2. `startOptimization`

The `startOptimization` method initiates the optimization process using the specified method and control parameters.

### Example Usage

Here is an example of how to configure and call the `BMLMOptimization` class, including the evaluation of initial values and starting the optimization:
  
```{r echo = TRUE, eval=FALSE}
# Initialize the BMLMOptimization object
bmlmOpt <- BMLMOptimization$new(
  projectConfiguration = myProjectConfig,
  runName = "MyOptimizationRun",
  scenarioList = myScenarios,
  dataObserved = myObservedData,
  seed = 1234
)

# Evaluate initial values
bmlmOpt$evaluateInitialValues()

# Start the optimization process
bmlmOpt$startOptimization(
  method = "BFGS",
  control = list(maxit = 1000),
  scalingMethod = SCALINGMETHOD$logsig,
  lastStatusSavingIntervalInSecs = 60,
  withInternalOptimization = TRUE
)
```
  

In this example:
  - The `BMLMOptimization` object is created with the necessary project configuration and data.
- Initial values are evaluated to ensure the model is set up correctly.
- The optimization is started using the "BFGS" method with specific control settings.

## Scaling Methods

Proper scaling of parameters is essential to ensure that they are on a similar scale, which helps prevent suboptimal solutions during optimization. The package offers flexible scaling methods, including linear and logarithmic scaling, allowing users to normalize parameter values based on their defined properties: `minValue`, `maxValue`, and `scaling`.

### Step 1: Scaling Parameters to a Value Between 0 and 1

The first step in the scaling process is to transform each parameter value to a normalized range between 0 and 1. This is achieved using one of two methods: linear scaling or logarithmic scaling.

#### Linear Scaling

In linear scaling, the transformation is performed using the following formula:
  
  $$ 
  \text{scaledValue} = \frac{\text{value} - \text{minValue}}{\text{maxValue} - \text{minValue}} 
$$
  
  This formula adjusts the parameter value based on its minimum and maximum bounds, ensuring that the output falls within the range of 0 (when the value equals the minimum) and 1 (when the value equals the maximum).

#### Logarithmic Scaling

Logarithmic scaling is particularly useful when the parameter values span several orders of magnitude. The transformation is defined as:
  
  $$ 
  \text{scaledValue} = \frac{\log(\text{value}) - \log(\text{minValue})}{\log(\text{maxValue}) - \log(\text{minValue})} 
$$
  
  This formula uses the natural logarithm to scale the parameters, also ensuring that the output is constrained between 0 and 1, provided that the values are within the valid range.

### Step 2: Handling Boundaries

After scaling the parameters to a value between 0 and 1, the next step is to manage the boundaries of the parameter space. This is achieved through two scaling options: "`r SCALINGMETHOD$hardBounds`" and "`r SCALINGMETHOD$logsig`" which can be set as input variables in the `BMLMOptimization$startOptimization` function.

#### Scaling method "`r SCALINGMETHOD$hardBounds`"

In the bounded scaling method, parameters are constrained to remain within the specified minimum and maximum values. If a parameter value exceeds these bounds during optimization, it is treated as NA (not available), indicating that the parameter is out of the allowable range.

#### Scaling method "`r SCALINGMETHOD$logsig`"

The `r SCALINGMETHOD$logsig` scaling method allows for a broader range of parameter values. This method transforms the parameters using a logistic function, which maps any real-valued number to a value between 0 and 1. This approach is particularly useful for parameters that may take on extreme values.

## Internal Optimization in Parameter Fitting

### withInternalOptimization

The `withInternalOptimization` parameter is a logical flag used in the `optimizeParameters` function to control whether an internal optimization step should be performed after each iteration of the main optimization process.

#### Purpose

In a hierarchical Bayesian model, the optimization of model parameters can be complex and computationally intensive, especially when the model involves multiple parameter types, including hyperparameters that define the distribution of individual parameters. The `withInternalOptimization` parameter allows for a more refined optimization approach by enabling the re-fitting of hyperparameters during the optimization process.

#### Functionality

- **When `withInternalOptimization` is set to `TRUE`:**
  - After evaluating the time profiles and calculating the log-likelihood for the current parameter values, an internal optimization step is triggered.
- This internal optimization focuses specifically on hyperparameters, allowing them to be adjusted based on the most recent estimates of the model parameters.
- The internal optimization is executed using the `runInternalOptimization` function, which optimizes the hyperparameters to maximize the log-likelihood.
- The results of the internal optimization are then used to update the parameter values, potentially leading to improved estimates in subsequent iterations of the main optimization.

- **When `withInternalOptimization` is set to `FALSE`:**
  - The optimization process proceeds without the additional internal optimization step. The model parameters are optimized based solely on the current estimates without adjusting hyperparameters during the optimization iterations.

#### Benefits

Enabling internal optimization can lead to:
  - Improved convergence of the optimization algorithm by allowing hyperparameters to adapt to the current parameter estimates.
- Potentially better fitting of the model to the data, as hyperparameters are refined continuously throughout the optimization process.



## Conclusion

The `BMLMOptimization` class provides a robust framework for configuring and executing the optimization of parameters in Bayesian hierarchical models. By understanding and effectively utilizing the `evaluateInitialValues` and `startOptimization` methods, along with the various scaling methods, users can enhance the performance and accuracy of their optimization tasks.

For more information on the specific functions and their usage, please refer to the package documentation.

