---
title: "Maximum likelihood estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{loglikelihood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(ospsuite.bmlm)
```


## Maximum likelihood estimation

The goal of maximum likelihood estimation is to find the values of the model parameters that maximize the likelihood function

$$ \hat {\theta } = \underset {\theta \in \Theta }{\operatorname {arg\;max}\mathcal {L(\theta,\mathbf {y},\mathbf{p})} }$$

$\theta$ is the vector of all adjustable parameters with in the maximum likelihood estimation

For a global parameter identification parameter it is one scalar $\theta_k$, k index of identification parameter 

For an individual identification parameter with hyper parameter  $\theta_k$ is a vector consisting of one scalar for each individual and the distribution parameters for the hyperparameter 

For all elements of $\theta$ which correspond to a global identification parameter or a distribution parameter of a hyperparameter belongs a prior distribution: $\mathbf{p}$

$\mathbf{y}$ is the vector of observed data over all individuals and observer.


To use this as objective function in an optimization algorithm it is converted :

$$ \hat {\theta } = \underset {\theta \in \Theta }{\operatorname {arg\;min}\mathcal {log(L(\theta,\mathbf {y},\mathbf{p}))} }$$


## Weighting of Data

For each data point a weighting is defined: $w_{i,j}(t)$ i individual, j observer, t time.

## construction of loglikelihood


$$ log(\mathcal{L(\theta,\mathbf {y})}) = log(\mathcal{L}_t(\theta,\mathbf {y}))  + log(\mathcal{L}_h(\theta))  + log(\mathcal{L}_p(\theta,\mathbf{p}))$$ 

- $\mathcal{L}_t$ Likelihood observed data given simulated time profiles
- $\mathcal{L}_h$ Likelihood of hyper parameter
- $\mathcal{L}_p$ Likelihood of parameter estimates given prior distribution



## $log(\mathcal{L}_t(\theta,\mathbf {y}))$ likelihood for timeprofiles


$$ log(\mathcal{L}_t(\theta,\mathbf {y})) =  \sum_{i,j}(log(\mathcal{L}u_{i,j}(\theta,\mathbf {y}))   + log(\mathcal{L}c_{i,j}(\theta,\mathbf {y})))  $$

- $i$ individual
- $j$ observer (Metabolite, Parent, fraction excreted Plasma, IV, PO,...)
- $\mathcal{L}u(\theta,\mathbf {y})$ Likelihood for uncensored  data points(values above lloq)
- $\mathcal{L}c(\theta,\mathbf {y})$ Likelihood for censored (values below lloq)




### Definitions per error model

description  | | additive | multiplicative | logarithmic
----------- | -------- | -------- | -------------- | -----------
residuals |  $\mathbf{r}_{i,j}(t)$ | $(\mathbf{y}_{i,j}(t) - f_{i,j}(\theta,t))$ | $(\mathbf{y}_{i,j}(t) - f_{i,j}(\theta,t_u))$ | $(log(\mathbf{y}_{i,j}(t)) - log(f_{i,j}(\theta,t)))$
residuals censored | $\mathbf{r}_{LLOQ,i,j}(t)$ | $(lloq_{j}(t) - f_{i,j}(\theta,t))$ | $(lloq_{j}(t) - f_{i,j}(\theta,t_u))$ | $(log(lloq_{j}(t)) - log(f_{i,j}(\theta,t)))$
residuals to physiological possible lower bound LB (typical = 0)| $\mathbf{r}_{LB,i,j}(t)$ | $(LB - f_{i,j}(\theta,t))$ | $(LB - f_{i,j}(\theta,t_u))$ | $(log(LB) - log(f_{i,j}(\theta,t)))$
model error vector|  $\epsilon_{i,j}(t)$ | $\epsilon_j$ | $\epsilon_j*f_{i,j}(\theta,t)$ | $\epsilon_j$
normalization vector|  $\epsilon N_{i,j}(t)$ | $\epsilon_j$ | $\epsilon_j$ | $\epsilon_j*d_{i,j}(t)$

- $\mathbf{y}_{i,j}(t)$ observed data point for individual i and output j
- $f_{i,j}(\theta,t)$ simulated data point for individual i and output j
- $lloq_{j}(t)$ lower limits of quantification
- $\epsilon_j$ model error by observer

<u>uncensored data points</u>

   $$log(\mathcal{L}u_{i,j}(\theta,\mathbf {y})) =  -\frac{1}{2}\sum_{t_u}w_{i,j}(t_u)\left( \frac{-(\mathbf{r}_{i,j}(t_u))^2.}{\epsilon_{i,j}(t_u)^2} - log(2\pi * \epsilon N_{i,j}(t_u)^2)\right)$$


<u>censored data points</u>

$$ log(\mathcal{L}c_{i,j}(\theta,\mathbf {y})) = \sum_{t_c}w_{i,j}(t_u)\left(log\left(\Phi\left(\frac{\mathbf{r}_{LLOQ,i,j}(t_c)}{\epsilon_{i,j}(t_c)}\right) - \Phi  \left(\frac{\mathbf{r}_{LB,i,j}(t_c)}{\epsilon_{i,j}(t_c)}\right)\right) -log\left(1-\Phi  \left(\frac{\mathbf{r}_{LB,i,j}(t_c)}{\epsilon_{i,j}(t_c)}\right)\right)\right)$$    
  
$$ \Phi(x) = \frac{1}{\sqrt{2*\pi}}\int_{-\inf}^xe^{-\frac{x^2}{2}} $$

  

## $log(\mathcal{L}_h(\theta))$ likelihood for hyperparameter 

$$ log(\mathcal{L}_h(\theta))  = \sum_k log(\mathcal{L}_{h,k}(\theta)) = \sum_{k \in K_h} log(\phi(\theta_{k}))$$

- $K_h$ parameter parameter with hyperparameter
- $\phi(\theta_k)$ distribution for hyperparameter  
- $\theta_{k}$ individual parameter values and distribution parameter values for hyperparameter k

available distributions:

- normal
- lognormal

easily extendable to all distributions available in package stats: (https://rdrr.io/r/stats/Distributions.html)


example for a multimodal normal distributed hyper parameter p:

$$log(\mathcal{L}_{h,k}(\theta))  = \sum_g\left(-\frac{1}{2\sigma_{k,g}^2}\sum_{i \in g}(value_{k,i} -\mu_{k,g})^2 -\frac{1}{2}log(2\pi*\sigma_{k,g}^2)\right) $$
- $i \in g$ individuals belonging to group g
- $value_{k,i}$ individual value for parameter k
- $\mu_{k,g}$ hyper parameter mean of mode g   
- $\sigma_{k,g}$ hyper parameter variance  of mode g 



## $log(\mathcal{L}_p(\theta,\mathbf{p}))$ likelihood for prior

$$ log(\mathcal{L}_p(\theta,\mathbf{p}))  = \sum_{k \in K_p} log(\mathcal{L}_{p,k}(\theta,\mathbf{p})) = \sum_k log(\Pi(\theta_{k},\mathbf{p}))$$

- $K_p$ elements of $\theta$ with prior
- $\Pi(\theta_k)$ distribution of prior

available distributions:

- flat:  $log(\mathcal{L}_{p,k})$ = 0  
- normal
- lognormal


